Visualizing Project Evolution Through Abstract Syntax Tree Analysis

### 2016-06-20 paper/document.tex
Add colour perception cite.

Source code is often represented as text in a source code file. This text is then parsed by a compiler or interpreter with the intent of eventually executing the code represented by the source text. While programmers may type in individual characters, their intention is at a much higher level. Their intent is written in tokens that represent statements, identifiers, and types that are used to solve their problems. Programmers solve problems in a way that is readable to other developers and executable by a computer. The programmer rarely operates at the character level, or even the line level, so why should our analyses of source code operate at granularities unfamiliar to developers?
The plain text nature of software does not stop at source code level. The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which store the differences in lines of text in a file between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Most hosting services for Git repositories provide some basic tools for analyzing projects over time~\cite{github-graphs,bitbucket-graphs}; however, as we will show, these tools often lack the kind of detail to gather insightful analysis. As with source code, commits operate at the text level. While commits may store individual changes as lines of code, they imply a change in \emph{structure} of the software.  By visualizing lines changed, or files changed, the current state-of-the-art for software project visualization rarely capture the \emph{meaning} behind each commit; lines of code alone does not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed does not necessarily reflect meaningful changes to a repository~\cite{robles2014}.

Consider GitHub graphs~\cite{github-graphs}, visualizations that are easily available to open source software developers. GitHub's graphs use line and commit information exclusively. In the case of the ANTRL4 project (Figure~\ref{fig:GH_code_freq}), this code frequency chart creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to crop the graph to be able to show details of other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by a number of reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps a massive refactoring effort~\cite{Hindle:2008:LCT:1370750.1370773}.

The goal of a software evolution visualization is to provide an alternate view that provides insight into the development of the software that is otherwise unavailable. By visualizing changes to lines and files, we are tracking the changes in text---not software. What would be more meaningful are the structural changes in code. These are changes in variable declarations, invocations of methods, and usage of types.

In this paper, we propose a method that analyses the changes in type usage and method invocations over time. Using commits, we are able to keep track of \emph{who} is changing what types. We then present this data using simple visualizations. We claim that type usage and method invocation offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. For simplicity, we analyze only Java projects, as it allowed us to quickly develop and test our software. However, since the approach described in this paper can be generally applied to any statically-typed language, it can be extended to other languages. Our goal is to describe a methodology for obtaining richer, more meaningful data from source code repositories which can be used in more insightful analyses.

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualize changes to types used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in Section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage over time from repositories? \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

Analyzing source code is a popular research topic and Java is often chosen as the subject of study due to its popularity. Grechanik \textit{et al.}~\cite{Grechanik:2010:EIL:1852786.1852801} examined the structure of Java programs mined from 2080 programs. This paper examined the breakdown of syntactic structures in open source repositories, however it does not consider per author statistics. Parnin \textit{et al.}~\cite{Parnin:2011:JGA:1985441.1985446} mined repositories to see how Java generics have been used in open source projects and found that generic usages were often introduced by a single developer in a project and were primarily being used for collecting and traversing lists of objects.

@@

The focus of these papers is on how programming languages and features are used by looking at many repositories. Java is a statically-typed language, yet there is no focus specifically on how types are used throughout a project. Type data may provide additional information on how developers use languages but this has yet to be explored.

@@

A variety of visualizations tools have been created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421,voinea2006} over time.

@@

Some tools examine the evolution of projects to learn other information about them. Alcocer \textit{et al.}~\cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined.

@@

The basic tools easily available to any software developer are the GitHub graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the type of code which was committed. It does not matter what language is used or if errors exist in the commit, GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of what is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

Of the services we have surveyed, none of the tools visualize project evolution using granular type analysis per author via commit-by-commit abstract syntax tree differencing.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section~\ref{sec:viz}.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by invoking \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure~\ref{fig:commits}. Since we compare each commit to its parent as reported by Git, we are able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure~\ref{fig:commits}, it indicates a merge commit in Git. Merge commits contain the changes made in each parent commit combined. Therefore, we ignored merge commits since all changes in the merge should already have been accounted for in previous commits. Thus, any branch that has been merged into the primary branch would have been analyzed and included by TypeV.

@@

@@

@@

\subsection{Views}

TypeV provides a variety of views to help organize the data and provide different levels of detail. The overview displays several rows of stacked bar graphs which show changes to a type or method over a period of time in a project (Figure~\ref{fig:main-view}). The purpose of displaying stacked bar charts is that they are a familiar visualization, thus are somewhat easy to read, and can succinctly display the relative amounts of additions and deletions per each time period. The light blue \additionsbox{10pt} and dark red \deletionsbox{10pt} portions account for the number of additions and deletions respectively, in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). These colours were chosen as they are high contrast, and are distinct to those experiencing red-green colour blindness.

The height of each bar shows the relative number of changes to a type. The width of each bar is fixed, and represents a time period for binning commits that can be configured to an hour, day, week, or month's worth of commits per each bar.  A bar that completely fills its row vertically is the period with the largest number of changes to that type in the visible date range. The height of the other bars in the row are in proportion to the maximum number of changes. The bar height is calculated by multiplying the maximum bar height by the number of changes during that time period divided by the maximum number of changes attested for this type. The rows are sorted so that the graph for the most used types or methods appear the top and are in descending order of use.

One can interact with the overview by hovering and clicking on the bars. When a user mouses over a specific bar in the graph they are given summary information for the type and time period at that location. The information includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message. By adding interaction, the user is able to explore the overview, find the specific information they are interested and drill-down at their leisure. Although our tool extracts a great deal of information, by using interaction, users can gradual reveal more information as they need it.

To allow for a more detailed analysis of a specific component of the repository we added some basic filtering tools. Users can filter by type name; they can view changes that fall within a specific start and end date; users can also choose to see only declarations of types, types used in declarations and invocations, or only invocations. One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different users in the \texttt{git log}. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple users at a time, thus allowing the user to view changes from an individual or group of developers.

Knowing type and author information enables other kinds of visualizations. For example, Figure~\ref{fig:chord} is a chord diagram which shows connections between authors and types. The wider the arc around the circumference of the circle, the more often that entity is connected to other entities. For example, ``parrt'' is the top contributor, with changes covering a large number of types. Determining author and type information is made possible by AST differencing.

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{chord}
\caption{Chord diagram showing connections between authors (right and bottom) to types (left).}
\label{fig:chord}
\end{figure*}

@@

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{antlr-density}
\caption{Density plot of type and file coverage for commits. The graph shows what percentage of type and file coverage the majority of commits have. One can see that the majority of commits have a type and file coverage around 0\% to 20\%. We can also see a difference between these measurements on the higher end of the graph. File coverage seems to have a group of commits that had a file coverage between 20\% and 50\% whereas type coverage have a similar looking grouping but between 40\% and 60\%.}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{tika-density}
\caption{Density of type and file coverage for commits. Here we can see that commits did not have much file coverage between 25\% and 75\%. On the other hand there were a lot of commits that have type coverage around 50\%.}
\end{figure}

Is it worth going through the effort of computing commit-by-commit AST differences if we can just simply count changes to files instead? We define \emph{type coverage} as the number of types that have been touched out of the total number of types during either a certain time period or the entire project. \emph{File coverage} is the same measurement, but using the number of source files touched instead. File coverage can be trivially determined from an author's commit history; Type coverage, however, is uniquely provided by AST differencing, as we have described.

In Java, a type can be a primitive, provided by the language itself, or an object reference. Primitive types are those that are built into the language such as an \texttt{int} or \texttt{char}. Object references point to instances of classes. Java's syntax encourages the creation of classes, thus we expect substantial Java projects to define many domain-specific types. Additionally, Java places the constraint that each source file must define one class. When writing or contributing to a project, the author's choice of which part of the program to edit will in turn affect what types they end up working with --- that is, it will affect the author's type coverage. Similarly, the set of files that an author works with affects that author's file coverage.

To determine if type coverage yields different information than file coverage, we analyzed four open source Java projects: ANTLR4, Apache BookKeeper, Apache Curator and Apache Tika. Three of the four projects were collected from the Apache foundation, as they host a large number of well-curated open source Java projects. A summary of the projects studied is given in Table~\ref{tab:project-summary}. We counted the file coverage and type coverage on a commit-by-commit basis for each author. That is, each datum represents a commit with three dimensions: file coverage and type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the statistics measured. In general, type coverage and file coverage are strongly correlated, both using Pearson (linear) correlation, and Spearman (rank) correlation. As type coverage and file coverage are both cumulative measures, positive correlation is expected. However, even when the distributions are significantly different (as with ANTLR and Curator), the correlation is still strongly positive, but not 100\%. This means that when the coverage for types and files of a commit increase, the relationship is not necessarily one-to-one. This is remarkable in a language like Java, as classes and files are generally in a one-to-one relationship.

@@

\subsection{ANTLR4}

@@

The purpose of \emph{deprecating} a software features is to give developers time to change projects to follow a new or better standard without breaking the code. When a piece of code becomes deprecated, continuing its use may lead to design flaws or even security vulnerabilities. For example, in Apache Nifi, an earlier version of the software used a compromised key-derivation function; hence, this class was deprecated, but left in for compatibility with older versions of the software~\cite{nifi}. Viewing changes in terms of types can show when usage of deprecated classes are added or removed from a project as well as how many remain over time. Given the time at which a feature was deprecated, seeing the progress of replacement has some important implications. In the case of a security flaw, that time could indicated times at which systems or users were vulnerable. If a feature is marked for removal, a project using a deprecated type has a limited amount of time to make the changes. Removing all instances of deprecated features could also mean a project can be upgraded to use new tools or features that were not possible while the deprecated features were present.

@@

@@

We only look at the master branch of a repository and ignore other branches; therefore, we might miss some of the work done by an author in another branch. As well, we ignore merge commits such that we do not double count changes; however, if there is a merge conflict, it is possible for the developer to makes changes. Changes made during merge conflict resolution are missed by TypeV. We have made the assumptions that if this case happens then the changes are minor.

@@

TypeV only extracts declaration of types and invocation of methods; it would be useful to also view what classes changes are being made to. Enabling this ability would also enable analysis of self-invocations.

@@

\textbf{RQ1:} How can we extract type and method usage over time from repositories?

@@

>>>>

Source code is often represented as text in a source code file. This text is then parsed by a compiler or interpreter with the intent of eventually executing the code represented by the source text. While programmers may type in individual characters, their intention is at a much higher level. Their intent is written in tokens that represent statements, identifiers, and types that are used to solve their problems. Programmers solve problems in a way that is readable to other developers and executable by a computer. The programmer rarely operates at the character level, or even the line level, so why should our visualizations of source code operate at granularities unfamiliar to developers?

The plain text nature of software does not stop at source code level. The history and evolution of a software project is often tracked by version control systems, such as Git, in a programming language agnostic way. Updates to source code, documentation, and tests are recorded in the form of commits, which store the differences in lines of text in a file between each subsequent revision of the software. Version control systems become vital for large projects since individual changes would be impossible to track manually. Large projects have thousands of individual changes made by many different developers. With more changes to peruse, it becomes more difficult to see how a project is evolving over time since there is so much information to look through. Hosting services for Git repositories often provide some basic tools for analyzing projects over time~\cite{github-graphs,bitbucket-graphs}; however, as we will show, these tools often lack the kind of detail to gather insightful analysis. As with source code, commits operate at the text level. While commits may store individual changes as lines of code, they imply a change in the \emph{structure} of software.  By visualizing lines changed, or files changed, many current tools for software projects rarely capture the \emph{meaning} behind each commit; lines of code alone do not describe how a repository changed at a particular point in time. Thus, information regarding contributions from developers is often misleading since the number of commits and lines of code changed do not necessarily reflect meaningful changes to a code repository~\cite{robles2014}. In this paper, we will show that abstract syntax trees (ASTs) can be used to create \emph{semantically-aware} software evolution visualizations.

Consider GitHub graphs~\cite{github-graphs}: visualizations that are easily available to the many software developers using GitHub to host their software repositories. GitHub's graphs use line and commit information exclusively. In the case of the ANTRL4 project (Figure~\ref{fig:GH_code_freq}), this code frequency chart creates more questions than it answers. What happened at those big spikes of line changes in the graph? One of the spikes is so large, we had to crop the image in order to show the magnitude of the other changes. How did these large changes affect the code base? Who committed these large changes? Was it a team of contributors or one particularly prolific programmer? Plotting the number of lines changed may be misleading. A huge spike in additions may be caused by a number of reasons, such as the inclusion of automatically-generated documentation, the inclusion of an external library, or perhaps a massive refactoring effort~\cite{Hindle:2008:LCT:1370750.1370773}.

The goal of a software evolution visualization is to provide an alternate view that provides insight into the development of the software that is otherwise unavailable. By visualizing changes to lines and files, one is tracking the changes in text---not software. Instead, we promote \emph{semantically-aware visualizations} that visualize the structural changes in code. In our case these are changes in variable declarations, invocations of methods, and usage of types. Others have explored semantically-aware visualization~\cite{sourcevis,6980224,Kim:2009:DRS:1555001.1555046,7332410}, however, we show that this can be done with ASTs.

In this paper, we propose a semantically-aware visualization method that tracks the changes in type usage and method invocations over time. Using commits, we are able to keep track of \emph{who} is using what types. We then present this data using time-proven visualizations. We claim that type usage and method invocation offers more insight than line-based, file-based, and commit-based methods of visualization and analysis. For simplicity, we analyze only Java projects, as it allowed us to quickly develop and test our software. However, since the approach described in this paper can be generally applied to any statically-typed language, it can be adapted to other programming languages. Our goal is to describe a methodology for obtaining richer, more meaningful data from source code repositories which can be used for more insightful analyses.

In the following sections of the paper we will describe our tool, TypeV, that we developed to compute and visualize changes to types by author used in a project over time. TypeV is able to extract type information from a repository by computing commit-by-commit abstract syntax tree differences, which we describe in Section~\ref{sec:methodology}. In this paper, we answer the following questions: \\ \\
\textbf{RQ1:} How can we extract type and method usage from repositories over time? \\ \\
\textbf{RQ2:} How does type coverage compare to file coverage?
\\ \\
\textbf{RQ3:} Does the type information give any interesting or useful insight into the project?

@@

Analyzing source code is a popular research topic and Java is often chosen as the subject of study due to its popularity and abundance of tools. Grechanik \textit{et al.}~\cite{Grechanik:2010:EIL:1852786.1852801} examined the structure of Java programs mined from 2080 programs. This paper examined the breakdown of syntactic structures in open source repositories, however it does not consider per author statistics. Parnin \textit{et al.}~\cite{Parnin:2011:JGA:1985441.1985446} mined repositories to see how Java generics have been used in open source projects and found that generic usages were often introduced by a single developer in a project and were primarily being used for collecting and traversing lists of objects.

@@

The focus of these papers was on how programming languages and features are used by looking at many repositories. Java is a statically-typed language---its types are checked at compile time---yet there is no focus specifically on how types are used throughout a project. Type data may provide additional information on how developers use languages but this has yet to be explored.

@@

A variety of visualizations tools have been created in order to help with understanding the changes that get made to software over time. Wu \textit{et al.}~\cite{wu2004a} explored punctuated changes to software repositories with a tool that generates an Evolution Spectrograph. In this visualization, each file has a row in the chart and each cell represents a section of time where an incoming or outgoing dependency can be changed. The colour is determined by the frequency of change. This work was continued to show file directories and display the cardinality of a developer's commits in Wu \textit{et al.}~\cite{wu2004}. The cardinality of a commit is the number of files or subsystems it affects in a system, indicating the scope of the change. Other tools also look at changes in code dependencies~\cite{6980224} or files~\cite{6650525, 7332421, voinea2006} over time.

@@

Some tools examine the evolution of projects to learn other information about them. Alcocer \textit{et al.}~\cite{6650523} create a Performance Evolution Blueprint, which displays the changes in performance over time as a project evolves. Ogawa and Ma created code\_swarm~\cite{codeswarm}, a generative art visualization which presents the progressive evolution of file changes made by authors in source control repository over time. Gource~\cite{gource} is a similar visualization, showing an animated force-directed tree of the file hierarchy as authors make changes to files in the source code repository over time. Ens \textit{et al.} created ChronoTwigger to visualize the co-evolution of source and test files~\cite{6980223}. This allowed testing practices to be more closely examined. Anslow \textit{et al.}~\cite{sourcevis} created SourceVis, a suite of software visualizations including visualizations for semantically-aware project evolution. SourceVis displays the evolution of Java source code metrics such as number of packages, classes, and methods, and fields.

@@

Developers using the massively popular GitHub website for their Git repository hosting instantly have access to basic history visualizations, known as GitHub Graphs (see Figure~\ref{fig:GH_code_freq})~\cite{github-graphs}. GitHub Graphs tracks contributions based only on the number of commits and number of lines changed. This has the advantage of not relying on the of code which was committed. It does not matter what language is used or if errors exist in the commit, GitHub will simply output the number of line additions and deletions. The issue with this type of analysis is that it does not give a clear idea of what is being changed, only that changes have occurred. This data may be lacking, or even misleading which contributes to the need for better tools.

Of the services we have surveyed, none of the tools that visualize project evolution address semantic history extraction using granular type analysis per author via commit-by-commit abstract syntax tree differencing.

@@

Given a Git repository, we want to extract per-author type changes over time. We accomplish this by computing the differences between abstract syntax trees of every consecutive pair of commits (Figure~\ref{fig:context}). This extracted data can then be used in a variety of visualizations as discussed in Section~\ref{sec:viz}. The reason for using ASTs is that they capture the semantics of the code far more than simple line counts.

@@

In order to gather all the changes for a given repository over time, we generated ASTs for each commit and compared the difference. We did this by running \texttt{git log} over the Git repository and listing all the commit IDs in chronological order (Figure~\ref{fig:commits}). For each commit, we collected the following information, as can be seen in Figure~\ref{fig:data-extracted}: the author's name, date of commit, commit ID (SHA), commit message, files the commit edited, and files in the repository at the time of the commit. For all the edited Java files we also compared the file at the time of the commit with its parent commit using AST differencing as described above. We achieved this by invoking \texttt{git difftool} and specifying the tool to be used as our AST differencing tool. One case where \texttt{git difftool} fails is when the commit has no parent, such as the initial commit. We were able to handle this by calling \texttt{git show} and saving the output to a temporary file to be used with the AST differencing tool. When calling \texttt{git show} instead of \texttt{git difftool} we treat everything as an addition, as in commit A in Figure~\ref{fig:commits}. Since we compare each commit to its parent as reported by Git, we are able to account for any branching that happened outside the primary branch of the repository (usually named \texttt{master}). If a commit has more than one parent, like commit D in Figure~\ref{fig:commits}, it indicates a merge commit in Git.

We deliberately chose to ignore merge commits. A merge commit has two cases to consider: The first case is when the parents of the commit can be automatically merged. In this case all the relevant changes are already recorded in the parent of the commits so we can safely ignore the merge commit~\cite{git:merge}. The second case occurs when the parents cannot be merged automatically: There is a merge conflict which requires manual resolution on the part of a developer. In this case the Git user will try to resolve the issue by modifying code. If we decide to include the merge commit we run into the issue of to how to compare the merge commit with its multiple parent commits and the possible modifications made to resolve the merge conflict. We could consider comparing the merge commit with only one parent commit at a time, however; we will end up counting all the changes already recorded in the other parent branch. Additionally, all changes made during the resolution of a conflict will be attributed to the author making the merge; this may exclude the author that actually wrote the code in the first place. The other option is to simply ignore the merge commit such that we do not count the changes done during a merge commit. We made the assumptions that merge conflicts occur infrequently and that changes done to resolve a merge conflict are usually minor. Therefore, we chose to ignore merge commits. Note that fast-forwards are not \emph{true merges}, hence they do not create merge commits~\cite{git:merge}; none of the previous cases involving merges occur.

@@

The major downside to using ASTs over a method like line count is that ASTs are more expensive to compute. Our experiments were run on a Intel i7 with 16GB of RAM and an SSD. Calculating the AST diffs varied from an hour or two for smaller projects to a couple of days for very large projects. For example ANTLR4 took about 8 hours while Apache BookKeeper only took a couple of hours. The good news for ongoing projects is that AST diffs only need to be computed for new commits; TypeV can resume calculating diffs and augmenting statistics at any time. New diffs can be appended to the already existing database of previous AST diffs.

@@

% CITE PAPERS THAT HAVE USED OUR APPROACH!
% AND SAY OUR VISUALIZATIONS ARE EFFECTIVE BY PROXY

% TODO: \cite{sillito2006}

@@

\subsection{Timeline View}
For our main demonstration of semantically-aware project evolution, we chose an interactive timeline view that displays the most popular types over time. The timeline displays several rows of stacked bar graphs which show changes to an individual type or method over a selectable period of time in a project (Figure~\ref{fig:main-view}). The light blue \additionsbox{10pt} and dark red \deletionsbox{10pt} portions account for the number of additions and deletions respectively, in a similar fashion to GitHub's code frequency graph (Figure~\ref{fig:GH_code_freq}). The height of each bar shows the relative number of changes to a type. The width of each bar is fixed, and represents a time period for binning commits that can be configured to an hour, day, week, or month's worth of commits per each bar. A bar that completely fills its row vertically is the period with the largest number of changes to that type in the visible date range. The height of the other bars in the row are in proportion to the maximum number of changes. The bar height is calculated by multiplying the maximum bar height by the number of changes during that time period divided by the maximum number of changes attested for this type.

One can interact with the overview by hovering and clicking on the bars. When a user mouses over a specific bar in the graph they are given summary information for the type and time period at that location. The information includes the fully-qualified name of the type, the number of commits, the number of authors, the number of additions, the number of deletions, and the start and end date of that time period. If the user clicks on the bar, they are given more detailed statistics which list the authors who made the commits and each commit message.

To allow for a more detailed analysis of a specific component of the repository we added filtering tools. Users can filter by type name, view changes that fall within a specific start and end date, and select whether to see only declarations of types, types used in declarations and invocations, or only invocations.

One issue with analyzing individual authors is that authors can commit from a variety of different computers which can have different emails and usernames attached to their Git configuration. This may result in one author appearing as several different authors in the \texttt{git log}. There is no perfect way to deal with this problem; however, we allow the user of our software to select and view multiple authors at a time, thus allowing the user to view changes from an individual or group of developers.

\subsection{Design Decisions}

The purpose of displaying stacked bar charts is that they are a familiar visualization. They can be viewed as an enhanced version of GitHub's code frequency graph~\cite{github-graphs}, with semantic data. We chose to create a similar visualization to GitHub to work with a visualization people may already be familiar with and evaluate our visualization tools' density of data.

The stacked bar graphs succinctly display the relative amounts of additions and deletions per each time period. This creates a dense view of the project's evolution. We chose to make the height of the bars relative to only the type itself to make it easy for a user to scan a single type and see when it was used modified. This makes tasks such finding a refactoring (see \ref{sec:antlr4}) very clear to see simply by browsing the line. We stacked the additions on top of the deletions such that users can immediately compare the relative frequency of changes of a single type throughout the chosen time span.

Our choice of colours was motivated to be high contrast and culturally indicative of their meaning. The particular colours chosen are high contrast, and are distinct to those experiencing protanomaly or deuteranomaly---red-green colour blindness. The red colour suggests a negative valence~\cite{moller2009}, hence suggesting deletions. The blue serves as a contrast to the red, indicating the opposite action (in this case, additions). Since

The rows are sorted so that the graph for the most used types or methods appear the top and are in descending order of use such that the most frequently used type is the first visible. The height allows users to see when certain types are being edited and ensures that highly used types do not overpower less used type making their changes less significant.

By adding interaction, the user is able to explore the timeline to find the specific information which they are interested in and drill-down at their leisure. Although our tool extracts a great deal of information, by using interaction, users can gradual reveal more information as they need it, rather than be bombarded with all of the information at once.

@@

% TODO: say all these other tools used file-based stuff:
% gource, github, code_swarm, others...?

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{antlr-density}
\caption{Density plot of type and file coverage for commits. The graph shows what percentage of type and file coverage the majority of commits have. One can see that the majority of commits have a type and file coverage around 0\% to 20\%. We can also see a difference between these measurements on the higher end of the graph. File coverage seems to have a group of commits that had a file coverage between 20\% and 50\% whereas type coverage have a similar looking grouping but between 40\% and 60\%.\label{fig:antlr4-density}}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\columnwidth]{tika-density}
\caption{Density of type and file coverage for commits. Here we can see that commits did not have much file coverage between 25\% and 75\%. On the other hand there were a lot of commits that have type coverage around 50\%.\label{fig:tika-density}}
\end{figure}

Is it worth the effort of computing commit-by-commit AST differences if we can just simply count changes to files? We define \emph{type coverage} as the number of types that have been touched out of the total number of types during either a certain time period or the entire project. \emph{File coverage} is the same measurement, but using the number of source files touched instead. File coverage can be trivially determined from an author's commit history. Type coverage, however, is uniquely provided by AST differencing, as we have described.

In Java, a type can be primitive, provided by the language itself, or an object reference. Primitive types are those that are built into the language such as an \texttt{int} or \texttt{char}. Object references point to instances of classes. Java's syntax encourages the creation of classes, thus we expect substantial Java projects to define many domain-specific types. Additionally, Java places the constraint that each source file must define one class. When writing or contributing to a project, the author's choice of which part of the program to edit will in turn affect what types they end up working with --- that is, it will affect the author's type coverage. Similarly, the set of files that an author works with affects that author's file coverage.

To determine if type coverage yields different information than file coverage, we analyzed four open source Java projects: ANTLR4, Apache BookKeeper, Apache Curator and Apache Tika. Three of the four projects were collected from the Apache foundation, as they host a large number of well-curated open source Java projects. A summary of the projects studied is given in Table~\ref{tab:project-summary}. We counted the file coverage and type coverage on a commit-by-commit basis for each author. That is, each datum represents a commit with three dimensions: file coverage, type coverage, and its commit date. Data is sorted in ascending order of commit date. Table~\ref{tab:summary} lists a summary of the statistics measured. In general, type coverage and file coverage are strongly correlated, both using Pearson (linear) correlation, and Spearman (rank) correlation. As type coverage and file coverage are both cumulative measures, positive correlation is expected. We displayed the density distributions for two different projects: ANTLR4 (Figure~\ref{fig:antlr4}) and Apache Tika (Figure~\ref{fig:tika-density}). However, even when the distributions are significantly different (as with ANTLR and Curator), the correlation is still strongly positive, but not 100\%. This means that when the coverage for types and files of a commit increase, the relationship is not necessarily one-to-one. This is remarkable in a language like Java, as classes and files are generally in a one-to-one relationship.

@@

\subsection{ANTLR4}\label{sec:antlr4}

@@

The purpose of \emph{deprecating} software features is to give developers time to change projects to follow a new or better standard without breaking the code. When a piece of code becomes deprecated, continuing its use may lead to design flaws or even security vulnerabilities. For example, in Apache Nifi, an earlier version of the software used a compromised key-derivation function; hence, this class was deprecated, but left in for compatibility with older versions of the software~\cite{nifi}. Viewing changes in terms of types can show when usage of deprecated classes are added or removed from a project as well as how many remain over time. Given the time at which a feature was deprecated, seeing the progress of replacement has some important implications. In the case of a security flaw, that time could indicated times at which systems or users were vulnerable. If a feature is marked for removal, a project using a deprecated type has a limited amount of time to make the changes. Removing all instances of deprecated features could also mean a project can be upgraded to use new tools or features that were not possible while the deprecated features were present.

@@

Having semantically aware evolutionary visualizations also gives users who are unfamiliar with the project at-a-glance information on the evolution of the project.

@@

Internal threats to validity include our treatment of merge conflicts. As mentioned before, we ignore all merges, so if there is a merge conflict, it is possible that a developer makes non-trivial changes. Changes made during merge conflict resolution are missed by TypeV. We have made the assumptions that if this case happens, the changes are minor.

@@

@@

\textbf{RQ1:} How can we extract type and method usage from repositories over time?

@@

Although calculating ASTs can be expensive, it is not computationally infeasible, and we benefit from semantically-aware code evolution visualizations that can be used to gain better insight into our software.
